---
title: "732A90: Lab 1"
author: "David Björelind, davbj395"
date: "10/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Include packages here


```

## Question 1: Be careful when comparing
```{r, echo=FALSE}
x1<- 1/3
x2<- 1/4

# First expression
if ( x1-x2 == 1/12 ) {
print ("Subtraction is correct")
} else {
print ("Subtraction is wrong")
}

# Second expression
x1 <- 1
x2 <- 1/2
if ( x1-x2 == 1/2 ) {
print ("Subtraction is correct")
} else {
print ("Subtraction is wrong")
}
```

The first expression tells us "Subtraction is wrong" and the second expression "Subtraction is correct". The first get it incorrect, because 1/3-1/4 cannot be represented in an exact way in binary. 1-1/2 can be represented correctly, which is why it gets the calculations correct.


Improvements:
use the comparing statement 'all.equal()' instead of using '==' will help this.

```{r}
options(digits=20)
1/3-1/4
1/12

x1<- 1/3
x2<- 1/4

# First expression again
if ( all.equal(x1-x2,1/12) ) {
print ("Subtraction is correct")
} else {
print ("Subtraction is wrong")
}
```
Now it gets it correct!

## Question 2: Derivative
```{r, echo=FALSE}
derivative = function(x, func, epsilon){
  return((func(x+epsilon)-func(x))/epsilon)
}
f = function(x){
  return(x)
}

epsilon = 10^-15

# Evaluating derivative
ders = c()
for (i in 1:10000){
  temp = derivative(i, f, epsilon)
  ders = c(ders, temp)
}

plot(x=1:10000, y=ders, type = 'l', main="Plot of derivatives", xlab="x-values", ylab="Derivatives: y'(x)")
```
The results are very surprising! The expected answer would be **1** for each value from 1-10000, f'(x) = 1. This is not observed however. The values quickly becomes 0, and this can be explained by underflow. **f(x+e)-f(x)** will produce a very small number. When **x** gets larger, the misrepresentation of **f(x+e)-f(x)** will grow and at some point it will be **0**.
Some experimentation: 

```{r}
(10+epsilon)-10
(15+epsilon)-15
(20+epsilon)-20
```

## Question 3: Variance
```{r, echo=FALSE}
# Variance function
myvar = function(x){
  n = length(x)
  ave = sum(x)/n
  temp = sum(x^2-ave^2)
  return(temp/(n-1))
}
# Generating numbers
xx = rnorm(n=10000, mean=10^8, sd=sqrt(1))

y = c()
normvar = c()
for (i in 2:10000){
  my_var = myvar(xx[1:i])
  vardif = my_var - var(xx[1:i])
  y = c(y, vardif)
  normvar = c(normvar, var(xx[1:i]))
}
# Plottning the results 
plot(x=2:10000, y=y, main="Plot of Y: difference between myvar(Xi) and var(Xi)", xlab = "Size of array used", ylab = "Y = myvar(Xi) - var(Xi)")
```
From the plot we can see that **myvar** differs from **var** +- 4 through the entire array. This difference is quite big since the real variance is **1** (from the random samples). We get the same error size range, no matter how many numbers are used for calculation. My function works well on samples with smaller values (10^3 for example). When increasing the subset on X with 1 sample, the variance from **myvar** increases or decreases drastically, while variance from **var** is stable. The problem here occurs when calculating **sum(x^2 - ave^2)** in **myvar**, because the actually small difference between **x^2 - ave^2** increases due to overflow which misrepresents these numbers. This is why I see such a big difference when increasing the sample size with 1.

Some examples follow below:
```{r}
myvar(xx[1:5001])
myvar(xx[1:5002])
myvar(xx[1:5003])
myvar(xx[1:5004])

```


```{r, echo=FALSE}
# Making a new var-function
myvar_new = function(x){
  mean = mean(x)
  n = length(x)
  var = 0
  for (i in 1:n){
    var = var + ((x[i]-mean)^2)/(n-1)
  }
  return(var)
}
# Generating numbers
xx = rnorm(n=10000, mean=10^8, sd=sqrt(1))

y = c()
normvar = c()
for (i in 2:10000){
  my_var = myvar_new(xx[1:i])
  vardif = my_var - var(xx[1:i])
  y = c(y, vardif)
}
# Plottning the results 
plot(x=2:10000, y=y, main="Plot of new variance function", xlab = "Size of array used", ylab = "Y = myvar(Xi) - var(Xi)")
```
Here, we see a much small spread from the real **var** function! The difference is that is intermittent steps is divided by **n**. Because of this, the numbers does not get too big. Thus, preventing overflow. However, we still get some overflow, especially for larger array sizes, but not in the same magnitude as before.

## Question 4: Binomial Coefficient
```{r, echo=FALSE}
# Writing all the functions
aa = function(n, k){
  return(prod(1:n) / (prod(1:k) * prod(1:(n-k))))
}
bb = function(n, k){
  return(prod((k+1):n) / (prod(1:(n-k))))
}
cc = function(n, k){
  return(prod(((k+1):n) / (1:(n-k))))
}
```

After experimenting some with different values I reach the following conclusions.
What happens when we move from A to C is than less and less expressions are being calculated via the **prod** function.
A: Cannot handle **k=0**. Answer should be **1** but returns **Inf**. This is because **prod(1:0) = 0**.
All: When **k=n** the result should be **1**, but all the functions return **Inf**. This occurs because they all contain **prod(1:(n-k))** which becomes **0** when **k=n**.

```{r, echo=FALSE}
# Playing around with large values for n and k first

a = c()
b = c()
c = c()
choose = c()
for (n in 2:500){
  k = floor(n/2)
  #k = n - 15
  a = c(a, aa(n, k))
  b = c(b, bb(n, k))
  c = c(c, cc(n, k))
  choose = c(choose, choose(n, k))
}
#log
a = log(a)
b = log(b)
c = log (c)
choose = log(choose)

plot(x=2:500, y=choose, main="Plot for choose() , A, B and C", col="black", type='l', ylim=c(-1, 350), xlab = "n (k = n/2)", ylab = "log-value")
lines(x=2:500, y=c, col="green", type='l')
lines(x=2:500, y=b, col="red", type='l')
lines(x=2:500, y=a, col="blue", type='l')
legend("topleft", legend=c("choose()", "c", "b", "a"), lwd=c(1,1,1,1),col=c("black", "green", "red", "blue"))

#plot(x=2:500, y=exp(choose), main="second plot", col="black", ylim=c(-1,10^120), type='l', xlab = "n (k = n/2)", ylab = "log-value")
#lines(x=2:500, y=exp(a), col="blue", type='l')
```
**Note**: I did log of all values to make then more manageable and that choose() is plotted behind C in the graph.

From reasoning above we suspect that A will overflow earlier than B due to more intermediate steps that needs to be calculated. With the same reasoning, B will overflow earlier than C.

From the plot we can see that both **A** and **B** stops showing at **n=180** and **n=280** respectively. This is explained by that larger n-values produce either **Inf** or **Nan**, which means that overflow is occurring. Choose() seems to have the same behavior as **C**.

Overflow is explained by the terms used in calculations. The term **prod(1:n)** makes **A** overflow first and **prod((k+1):n)** makes **B** overflow second. None of these term occurs in **C**, it avoids multiplying the largest numbers. Thus, preventing overflow.

### All code for this report:
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```