---
title: "732A90: Lab 1"
author: "David Björelind, davbj395"
date: "10/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Include packages here


```

## 1. Be careful when comparing
```{r, echo=FALSE}
x1<- 1/3
x2<- 1/4

# First expression
if ( x1-x2 == 1/12 ) {
print ("Subtraction is correct")
} else {
print ("Subtraction is wrong")
}

# Second expression
x1 <- 1
x2 <- 1/2
if ( x1-x2 == 1/2 ) {
print ("Subtraction is correct")
} else {
print ("Subtraction is wrong")
}
```

The first expression tells us "Subtraction is wrong" and the second expression "Subtraction is correct". The first get it incorrect, because 1/3-1/4 cannot be represented in an exact way in binary. 1-1/2 can be represented correctly, which is why it gets the calculations correct.


Improvements:
use the comparing statement 'all.equal()' instead of using '==' will help this.

```{r}
options(digits=20)
1/3-1/4
1/12

x1<- 1/3
x2<- 1/4

# First expression again
if ( all.equal(x1-x2,1/12) ) {
print ("Subtraction is correct")
} else {
print ("Subtraction is wrong")
}
```
Now it gets it correct!

## 2. Derivative
```{r, echo=FALSE}
derivative = function(x, func, epsilon){
  return((func(x+epsilon)-func(x))/epsilon)
}
f = function(x){
  return(x)
}

epsilon = 10^-15

# Evaluating derivative
ders = c()
for (i in 1:10000){
  temp = derivative(i, f, epsilon)
  ders = c(ders, temp)
}

plot(x=1:10000, y=ders, type = 'l')
```
The results are very surprising! The expected answer would be **1** for each value from 1-1000, f'(x) = 1. This is not observed however. The values quickly becomes 0, and this can be explained by underflow. **f(x+e)-f(x)** will produce a very small number. When **x** gets larger, the misrepresentation of **f(x+e)-f(x)** will grow and at some point it will be **0**.

```{r}
(10+epsilon)-10
(15+epsilon)-15
(20+epsilon)-20
```

## 3. Variance
```{r, echo=FALSE}
# Variance function
myvar = function(x){
  n = length(x)
  sum = sum(x)
  temp = x^2 - (sum/n)^2
  return(sum(temp)/(n-1))
}
# Generating numbers
xx = rnorm(n=10000, mean=10^8, sd=sqrt(1))

y = c()
for (i in 2:10000){
  var = myvar(xx[1:i])
  vardif = var - var(xx[1:i])
  y = c(y, vardif)
}
# Plottning the results 
plot(x=2:10000, y=y, main="Plot of y: difference between myvar(Xi) and var(Xi)")
```
When adding many large numbers, there becomes a problem.

## 4. Binomial Coefficient
```{r}

```

## Another chunk
```{r}

```

### All code for this report
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```