---
title: "732A90: Lab 4"
author: "David Björelind, davbj395"
date: "11/25/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Include packages here

```
## Question 1: Computations with Metropolis–Hastings

### 1.
```{r, echo=FALSE}
pdf = function(x){
  return(exp(-x)*x^5)
}

metro_hastings = function(start, niter){
  res = array(0, niter)
  res[1] = start
  for(i in 2:niter){
    y = rnorm(1, mean=res[i-1] , sd=1)#, log=TRUE)
    u = runif(1, 0, 1)
    alpha = min(1, (pdf(y)*dnorm(res[i-1], mean=y , sd=1, log=TRUE))/(pdf(res[i-1])*dnorm(y, mean=res[i-1] , sd=1, log=TRUE)) )
    if(u< alpha){
      res[i] = y
    } else {
      #print('alpha > u')
      res[i] = res[i-1]
    }
  }
  
  return(res)
}

samples = metro_hastings(20, 5000)
plot(samples, type='l', main='Samples from Metropolis Hastings: log-normal')
```

### 2.
```{r, echo=FALSE}
metro_hastings = function(start, niter){
  res = array(0, niter)
  res[1] = start
  for(i in 2:niter){
    y = rchisq(1, df=floor(res[i-1]))
    u = runif(1, 0, 1)
    alpha = min(1, (pdf(y)*dchisq(res[i-1], df=floor(y)))/(pdf(res[i-1])*dchisq(y, df=floor(res[i-1]))))
    if(u< alpha){
      res[i] = y
    } else {
      #print('alpha > u')
      res[i] = res[i-1]
    }
  }
  
  return(res)
}

samples = metro_hastings(20, 5000)
plot(samples, type='l', main='Samples from Metropolis Hastings: chi-2 distribution')
```

### 3.
There is a short burn-in period of about 20 iterations for both proposal distributions. The produces samples are varying from about 3 to 10 for both distributions. The mean and variance of the samples are also the same. Looking on the plots however, we can see some differences. Samples from the log-normal distribution is more smooth, and the samples from chi-2 distribution have higher (and lower) extreme values.


```{r, echo=FALSE}

```

### 4.
```{r, echo=FALSE}

```

### 5.
```{r, echo=FALSE}

```

### 6.
```{r, echo=FALSE}

```


## Another chunk
```{r}

```

### Include all code for this report
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```