---
title: "732A90: Lab 3"
author: "David Björelind, davbj395"
date: "11/18/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Include packages here
library(poweRlaw)

```

## Question 1: Stable distribution

### 1.
```{r, echo=FALSE}

func = function(x, c){
  res = c/(sqrt(2*pi))*exp(-c^2/(2*x))*x^(-3/2)*1
  return(res)
}

power_func = function(x, alpha, tmin){
  res = c()
  for(i in 1:length(x)){
    ones = if(x[i] >= tmin) 1 else 0
    f = (alpha-1)/tmin*(x[i]/tmin)^(-alpha)* ones
    res = c(res, f)
  }
  return(res)
}

xgrid = seq(0.001, 100, length.out = 200)

c = 6 # Will vary throughout the lab

# Majorizing constant
y = func(seq(0,100, length.out = 200), c)
major_const = max(y)

alpha = 2 # Pick a good value
tmin = 5 # Fixed for entire lab
f_val = func(xgrid, c)
power_val = power_func(xgrid, alpha, tmin)

grid = seq(0.001, 100, length.out = 10000)
vals = func(grid, c)
pos = sum(vals[1:(10000*(tmin/100))])/sum(vals) # Density between 0 to 1, meaning probability for value to occur there

fx = func(seq(tmin,100, length.out = 200), c)
fy = power_func(seq(tmin,100, length.out = 200), alpha, tmin)
#major_const = max(fx/fy)
major_const = 1

plot(x = xgrid, y = f_val, type='l', col="black", ylab = "y-values", ylim=c(0, 0.05), main="f and fp")
lines(x = xgrid, y = power_val*major_const, type='l', col="red")

```
One problem is that the two functions has different support. The power-law is supported on (Tmin, inf) while f(x) has support on (0, inf). This means that support for (0, Tmin) is missing from power-law. to account for this, a proposal density from a uniform distribution is used for (0, Tmin). Now, the proposal density has the support (0, Inf). **Alpha** is picked so that the shape of **fp** looks similar to **f**, see plot above. A higher number could have been used, but Alpha=2 works.

### 2.
```{r, echo=FALSE}

acc_rej = function(c, tmin, alpha, p){
  fx = func(seq(tmin,100, length.out = 200), c)
  fy = power_func(seq(tmin,100, length.out = 200), alpha, tmin)
  major_const2 = max(fx/fy)
  major_const1 = max(func(seq(0.001,100, length.out = 200), c))
  reject = 0
  er = p*(1-(1/major_const1)) + (1-p)*(1-(1/major_const2)) # Not correct
  
  while (1<2) {
    choice = rbinom(1,1,p)
    if (choice == 1){
      y = runif(1, 0, tmin)
      major_dens = 1
      major_const = major_const1
    
    } else{
      y = rplcon(n=1, xmin=tmin, alpha=alpha)
      major_dens = power_func(y, alpha, tmin)
      major_const = major_const2
      
    }
    u = runif(1, 0,1)
    if(u*major_dens*major_const <= func(y, c)){
      return(c(y, reject, er))
      
    } else {
      # Rejected!
      reject = reject+1
    }
  }
}

```
http://www.columbia.edu/~ks20/4703-Sigman/4703-07-Notes-ARM.pdf
### 3.
```{r, echo=FALSE}
n_samples = 1000

#mean = mean(samples[,1])
#var = var(samples[,1])
#rrate = sum(samples[,2])/n_samples
#cat('c = ', c, 'gives: \nMean rejection rate: ', rrate, '\nMean: ', mean, '\nVariance: ', var)

cs = c(3, 4, 5, 6, 7, 8)
for (j in cs){
  samples = matrix(nrow = n_samples, ncol = 3)
  for(i in 1:n_samples){
    sam = acc_rej(j, tmin, alpha, pos)
    #if (sam[1] >10000) sam[1]=0
    samples[i,] = sam
  }

  mean = mean(samples[,1])
  var = var(samples[,1])
  rrate = sum(samples[,2])/n_samples
  cat('\nc = ', j, 'gives: \nMean rejection rate: ', rrate, '\nMean: ', mean, '\nVariance: ', var)
  hist(samples[,1],breaks=(max(samples[,1])-min(samples[,1]))/1, freq = FALSE, xlim=c(0,100), main="Histogram of samples", xlab="")
  hist(samples[,2],breaks=100, freq = TRUE, main="Hisogram of rejections", xlab="")
}
```
It can be seen that when **c** increases, the mean and variance increases with it. Judging from the behavior of **c** on **f(x)**, as **c** decreases the higher the peak at the start becomes. A high **c**-value creates a low peak and a more smooth graph. 

Rejection rates...

## Question 2: Laplace distribution

### 1.
```{r, echo=FALSE}
# New function
# Inverse CDF method
inv_de = function(y){
  return(log(2*y))
}



```

### 2. 
```{r}
# Target density: N(0,1)
# Majorizing density: DE(0,1)


```



## Another chunk
```{r}

```

## Another chunk
```{r}

```
### Include all code for this report
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```